Executive Constraints and Civil Conflict Onset

### 2019-10-08 paper.Rmd
Research design, second draft

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: \today
documentclass: scrartcl
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
---

```{r include = F, results = "hide"}
knitr::opts_chunk$set(echo = F, eval = F)

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian multilevel logistic regression. In what follows, I first
describe the independent and dependentvariables, including the
operationalizations and data sources, and then delve into the
regression model before finally presenting the results.

@@

The dependent variable, *civil conflict onset*, was taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 (Cite) covering 1946-2018
where intra-state conflict is defined as a violent contest over a
government or territory between a state and non-state actor with a
minimum threshold of 25 battle deaths within a single country-year.
The data was merged based on country location for conflicts labeled as
internal or internationalized internal --- *i.e.* where
`type_of_conflict` was either 3 or 4. Multiple conflicts within the
same years were aggregated to form a binary variable coded as 1 to
denote the onset of at least one new civil conflict and 0 for all
other years, including years of ongoing conflict.^[Extrasystemic
conflicts, *i.e.* imperial or colonial conflicts, were excluded
largely due to lack of data. Ideally, such conflicts would be merged
based on the territory location of the conflict, however, for
colonies, for example, there is very little easily available economic
data and it would be undesirable to use the data for the colonizing
power due to the often wide discrepancies in development within
colonial empires. TODO: maybe a blurb on theory?]

@@

(TODO: Probably cite Sklar and/or O'Donnell here for more theoretical
justification of operationalization)

To measure formal executive constraints I use data from the Varieties
of Democracy (V-Dem) dataset v9.0 (cite V-Dem). V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, mostly academics as well as professionals
working in policy related fields (cite Methodology doc). Individual
ratings are aggregated into latent variable estimates using
hierarchical ordinal item response theory (IRT) models taking into
account rater reliability and varying rater thresholds (cite Pemstein
et al). ^[Thresholds relate to how the continuous latent trait for a
given variable is mapped on to the ordinal answer categories presented
on the survey. Due to differential item functioning, experts may
perceive the magnitude of difference between answer items
differently. For example, for the variable *v2elvotbuy*, election vote
buying, experts may vary in how they interpret the difference between
a 2 (Restricted Vote Buying) and a 3 (Almost No Vote Buying). These
two concepts, rater reliability and varying rater thresholds,
correspond respectively to the *discrimination* and *difficulty*
parameters in classical IRT models.] The result is over 220 lower
level indicators that are also combined through various aggregation
techniques to produce over 80 mid/high-level indices. Compared to
datasets such as Polity IV (cite), which has traditionally been
used in most analyses of civil conflict, V-Dem provides much more fine
grained, nuanced data, and most importantly, includes measures of
uncertainty propagated from the IRT models. This latter point is
crucial since ignoring measurement error, especially in models with
noisy data and small effect sizes as is often the case in conflict
studies, can bias estimates and lead to overconfident results (cite
Treier & Jackman; Kyle).

For the purposes of this paper, I re-estimate Lührmann *et al*'s
horizontal accountability index (*v2x_horacc*) in V-Dem with the
addition of several variables (cite). The *v2x_horacc* index aims to
capture the extent to which the executive is *de facto* constrained by
other government institutions --- more specifically, the legislature,
judiciary, constitution, and independent investigative bodies (cite pg
61) --- rather than the *de jure* powers of any one institutions
vis-à-vis the executive. It is estimated as a latent variable using a
structural equation model, thereby avoiding the arbitrary weighting of
linear additive indices.^[A theoretical argument could be made that
the oversight offered by each institution is a necessary condition and
that any index should therefore be additive with equal
weighting. However, this would be a normative measure capturing the
degree to which a country-year fulfills an ideal concept, rather than
measuring how constrained an executive actually is in practice.]

For the two principal institutions, the legislature and the judiciary,
to serve as effective constraints there needs to be both a *capacity*
and *willingness* to oversee the executive. To better capture the
former dimension, I include two additional legislative variables. In
addition to *v2lgqstexp*, the extent to which the legislature
questions executive branch officials, and *v2lginvstp*, the extent to
which the legislature investigates in practice, I add *v2lgfunds*, the
extent to which the legislature controls its own resources, and
*v2lgoppart*, the extent to which opposition parties are able to
exercise oversight. The remaining manifest variables are the
same. Namely, for the judicial component: *v2juhcind*, high court
independence; *v2juncind*, lower court independence; *v2juhccomp*, the
degree to which the government complies with high court decisions;
*v2jucomp*, the degree to which the government complies with lower
court decisions.

v2exrescon
v2lgotovst

How measured? FA model, include formula

Measurement error (cite?)

Special case for legislative variables

Imputation for v2lgotovst

DAG

\[ x_{jt}^\ast \sim \mathsf{Normal}(\lambda_j \theta_t, \psi_t) \]
\[ x_{jt}^{obs} \sim \mathsf{Normal}(x_{jt}^\ast, \tau_j^{obs}) \]

@@

@@

### Model

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *n* countries indexed
by $i = 1 \dotsc n$ and *m* years indexed by $t = 1 \dotsc m$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\pi_{it}$, is the inverse logistic such that $\pi_{it} \in [0, 1]$.
^[Technically, the inverse logistic transforms values to the open
interval, $(0, 1)$, but because of overflow/underflow in computer
calculations due to floating point math the endpoints can be
observed.]

The full regression specification is then as follows:

\[ y_{it} \sim \mathsf{Bernoulli}(\pi_{it}) \]
\[ \pi_{it} = \mathsf{logit^{-1}}(\alpha + X_{it-1} \beta + \sum_{k=1}^K \delta^k_{j_{it}}) \]

@@

Estimated using HMC by stan (footnote number of iterations, burnin, and convergence)

@@

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
documentclass: scrartcl
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
    html_document:
        css: assets/sakura.css
        smart: true
        toc: true
---

```{r include = F, results = "hide"}
# Start by setting up our R env
knitr::opts_chunk$set(echo = F, eval = F)

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian multilevel logistic regression. In what follows, I first
describe the independent and dependent variables, including the
operationalizations and data sources, and then delve into the
regression model before finally presenting the results.

@@

The dependent variable, *civil conflict onset*, was taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 (Cite) where intra-state
conflict is defined as a violent contest over a government or
territory between a state and non-state actor with a minimum threshold
of 25 battle deaths within a single country-year with a coverage from
1946 to 2018.  The data was merged based on country location for
conflicts labeled as internal or internationalized internal --- *i.e.*
where *type_of_conflict* was either 3 or 4. Multiple conflicts within
the same years were aggregated to form a binary variable coded as 1 to
denote the onset of at least one new civil conflict and 0 for all
other years, including years of ongoing conflict.^[Extrasystemic
conflicts, *i.e.* imperial or colonial conflicts, were excluded
largely due to lack of data. Ideally, such conflicts would be merged
based on the territory location of the conflict, however, for
colonies, for example, there is very little easily available economic
data and it would be undesirable to use the data for the colonizing
power due to the often wide discrepancies in development within
colonial empires. TODO: maybe a blurb on theory?]

@@

(TODO: Probably cite Sklar and/or O'Donnell here for more theoretical
justification of operationalization, maybe even define institutions as
well)

To measure formal executive constraints I use data from the Varieties
of Democracy (V-Dem) dataset v9.0 (cite V-Dem). V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, mostly academics as well as professionals
working in policy related fields (cite Methodology doc). Individual
ratings are aggregated into latent variable estimates using
hierarchical ordinal item response theory (IRT) models taking into
account rater reliability and varying rater thresholds (cite Pemstein
et al). ^[Thresholds relate to how the continuous latent trait for a
given variable is mapped on to the ordinal answer categories presented
on the survey. Due to differential item functioning, experts may
perceive the magnitude of difference between answer items
differently. For example, for the variable *v2elvotbuy*, election vote
buying, experts may vary in how they interpret the difference between
a 2 (Restricted Vote Buying) and a 3 (Almost No Vote Buying). These
two concepts, rater reliability and varying rater thresholds,
correspond respectively to the *discrimination* and *difficulty*
parameters in classical IRT models.] The result is over 220 lower
level indicators that are also combined through various aggregation
techniques to produce over 80 indices. Compared to datasets such as
Polity IV (cite), which has traditionally been used in most analyses
of civil conflict, V-Dem provides much more fine grained, nuanced
data, and most importantly, includes measures of uncertainty
propagated from the IRT models. This latter point is crucial since
ignoring measurement error, especially in models with noisy data and
small effect sizes as is often the case in conflict studies, can bias
estimates and lead to overconfident results (cite Treier & Jackman;
Kyle).

For the purposes of this paper, I re-estimate Lührmann *et al*'s
horizontal accountability index (*v2x_horacc*) in V-Dem with the
addition of several variables (cite). The *v2x_horacc* index aims to
capture the extent to which the executive is *de facto* constrained by
other government institutions --- more specifically, the legislature,
judiciary, constitutional norms, and independent investigative bodies
(cite pg 61) --- rather than the *de jure* powers of any one
institution *vis-à-vis* the executive. It is estimated as a latent
variable using a structural equation model thereby avoiding the
arbitrary weighting of linear additive indices.^[A theoretical
argument could be made that the oversight offered by each institution
is a necessary condition and that any index should therefore be
additive with equal weighting. However, this would be a normative
measure capturing the degree to which a country-year fulfills an ideal
concept, rather than measuring how constrained an executive actually
is in practice.]

For the legislature and the judiciary, the two principal government
branches most able to check the power of the executive, there needs to be
both a *capacity* and *willingness* for oversight. To better capture
the former dimension, I include two additional legislative
variables. Besides *v2lgqstexp*, legislature questions executive
branch officials, and *v2lginvstp*, legislature investigates in
practice, I add *v2lgfunds*, legislature controls its own resources,
and *v2lgoppart*, opposition parties able to exercise oversight. The
remaining manifest variables are unchanged. Namely, for the judicial
component: *v2juhcind*, high court independence; *v2juncind*, lower
court independence; *v2juhccomp*, the degree to which the government
complies with high court decisions; *v2jucomp*, the degree to which
the government complies with lower court decisions. Norms relating to
the status of the constitution are captured by *v2exrescon*, executive
respects constitution, and the final dimension, independent
investigative bodies, by *v2lgovost*, executive oversight by bodies
other than the legislature.

For the estimation of the latent index I largely follow the approach
taken by *v2x_horacc* with two notable exceptions. The first is that I
model the measurement error in the manifest variables using the
reported posterior standard deviations from V-Dem --- available as
_v2\*\_sd_. The second is that for computational reasons I allow the
judicial variables to impact the latent factor directly rather than
nested within a second latent factor (TODO: better justification). The
result is the following model specification.

For $m$ country-years indexed by $i \in I = \{1, \dotsc, n\}$ and $n$
manifest variables indexed by $p \in P = \{1, \dotsc, n\}$, let $Z \in
\mathbb{R}^{m \times n}$ be the matrix of manifest variables measured
with an observed degree of error, $Z^\ast \in \mathbb{R}^{m \times n}$
the matrix of latent true values, and $\theta$ the unidimensional
latent factor measuring executive constraints. Further, let $Q(Z_{\ast
p})$ be a property defined as true when $Z_{\ast p}$ denotes a
legislative variables and otherwise false.

I first model the measurement error in the manifest variables by
assuming that the observed values are drawn from a normal distribution
where the location parameters are the latent true values and the scale
the observed posterior standard deviations (cite McElreath).^[An even
more sensible approach would be to directly embed the individual IRT
models aggregating the manifest variables into the larger executive
constraints measurement model (cite Kyle). However, this prohibitively
increases the run time of the model.]
\[ Z_{ip} \sim \mathsf{Normal}(Z^\ast_{ip}, \tau_{ip}) \]

The latent values for the manifest variables are then modeled as
functions of the latent factor $\theta$. For the non-legislative
variables the mapping functions are identical to a standard factor
analytic approach.
\[
  Z^\ast_{ip} \sim \mathsf{Normal}(\gamma_p + \lambda_p \theta_i, \psi_p)
  \mbox{\quad if } p \in P \mid \neg Q(Z_{\ast p})
\]

where $\lambda$ is the variable specific factor loading, $\gamma$ the
population intercept, and $\psi$ the scale of the error terms.

In the case of the legislative variables there is the added issue of
missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves latent variables, minimal values cannot be imputed without
a nonlinear transformation to a bounded scale. I thus follow
*v2x_horacc* in mediating the effect of the latent factor on the
legislative variables by the presence of the legislature using a
dichotomized version of *v2lgbicam*, how many chambers does the
legislature contain, denoted by the Bernoulli distributed variable,
$w$, where country-years with a sitting legislature are given a $1$
and country-years where the legislature is missing or has been
suspended, $0$. The probability of a legislature being present,
$\pi_i$, is then a linear function of the latent factor, $\theta$,
transformed by the inverse logistic such that $\pi_i \in [0,
1]$.^[Technically, the inverse logistic transforms values to the open
interval, $(0, 1)$, but due to overflow/underflow in floating point
calculations by computers the endpoints can be observed.] The
legislative manifest variables are in turn treated as functions of the
linear predictor for $w$ (cite pg 63). The practical effect is that
where there is data for the legislative manifest variables, the latent
factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables.
\begin{align*}
  w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
  \pi_i & = \mathsf{logit}^{-1}(\xi_i) \\
  \xi_i & = \eta + \delta \theta_i \\
  Z^\ast_{ip} & \sim \mathsf{Normal}(\gamma_p + \lambda_p \xi_i, \psi_p)
    \mbox{\quad if } p \in P \mid Q(Z_{\ast p}) \land w_i = 1
\end{align*}

[^missing]: Missingness due to the lack of of a particular
institution, for example the suspension of high court, is an issue for
the other manifest variables as well. However, only the existence of
the legislature is coded by V-Dem as a separate variable. Furthermore,
practically speaking, none of the other manifest variables exhibit the
same systematic pattern of missingness in the dataset as the
legislative variables. Whereas there are 1608 country-years lacking a
legislature ($13.4\%$ of all post-1945 country-years), only 115 are
missing data for at least one of the judicial variables ($1\%$) and
only 58 are missing for *v2exrescon* ($0.5\%$). The variable
*v2lgotovst*, executive oversight by bodies other than the
legislature, is an outlier with 1619 country-years missing ($13.5\%$)
since there is an error in V-Dem's v9.0 dataset whereby it has been
cleaned according to when the legislature is missing. Thus, I impute
*v2lgotovst* within the larger model where the legislature is missing
by drawing missing values from a normal distribution with a mean and
standard deviation estimated from the observed *v2lgotovst* data. No
other country-years or variables are imputed. The rationale is that,
disregarding data processing issues, missingness in V-Dem can arise
either from a lack of expert coders or lack of information. The latter
scenario will most likely be correlated with country-years where there
is a crisis or significant political instability, such as coups or
civil war, where extreme values are more probable. Since it cannot be
assumed that the values are completely missing at random, the data
generative process for missingness cannot be modeled without
potentially introducing bias into the estimates.

Finally, for the both the legislative and non-legislative mapping
functions I assume local item independence such that the elements of
$\Psi$ are uncorrelated (cite Lord & Novick). Furthermore, for the
purposes of identifying the likelihood the scale of the latent factor,
$\theta$, is fixed to mean $0$ and standard deviation $1$ while the
slopes, $\Lambda$ and $\delta$, are constrained to be positive (cite
Jackman). The positive constraints are drawn from theory since all of
the manifest variables exhibit the same scale direction and should be
positively correlated with the latent factor. Similarly, the presence
of the legislature should be positively correlated with higher values
of the latent factor, *i.e.* greater executive constraints.

@@

In addition to the data on conflict onset and executive constraints, I
also include a number of standard controls drawn largely from previous
literature, the most important of which is state capacity. According
to the theory outlined earlier, the impact of executive constraints on
decreasing the probability of conflict onset should be conditional on
the relative capacity of the state. The lack of executive constraints
most likely increases the risk of civil conflict when state capacity
is low and the central government is willing to use repression, but
lacks the ability to fully suppress the opposition. To measure state
capacity I follow Fearon & Laitin in using GDP per capita as a proxy
for the state's administrative and military capacity by roughly
capturing the revenue potential available to the state (cite). As a
data source, I use the natural log of the *RGDP^e* variable, constant
price real GDP, from Penn World Tables v9.1 (1950-2017) to allow for
both cross-country and cross-temporal comparisons of GDP per capita
(cite)^[TODO: Expand on why Penn World Tables] and include an
interaction term between GDP per capita and the latent measure for
executive constraints to capture the conditional relationship.

@@

### Regression Model

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *n* countries indexed
by $i = 1 \dotsc n$ and *m* years indexed by $t = 1 \dotsc m$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\nu_{it}$, is the inverse logistic such that $\nu_{it} \in [0, 1]$.

The full regression specification is then as follows:
\begin{align*}
  y_{it} & \sim \mathsf{Bernoulli}(\nu_{it}) \\
  \nu_{it} & = \mathsf{logit^{-1}}(\alpha + X_{it-1} \beta + \sum_{k=1}^K \zeta^k_{j_{it}})
\end{align*}

@@

Measurement model:
\begin{align*}
  \theta & \sim \mathsf{Normal}(0, 1) \\
  \delta, \lambda & \sim \mathsf{LogNormal}(0, 1) \\
  \eta, \gamma & \sim \mathsf{Normal}(0, 5) \\
  \psi & \sim \mathsf{Weibull(2, 1)}
\end{align*}

Regression parameters:
\begin{align*}
  \beta & \sim \mathsf{Normal}(0, 2.5) \\
  \alpha & \sim \mathsf{Normal}(0, 5) \\
  \zeta^k & \sim \mathsf{Normal}(0, \sigma^k) \\
  \sigma & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}

@@

## Prior Predictive Check
